---
title: "Association of air pollution and cardiovascular disease rate in california counties"
author: "__Luqing Ren__"
always_allow_html: TRUE
output:
  html_document:
      toc: TRUE
      toc_float: TRUE
editor_options: 
  chunk_output_type: inline
---


```{r, include = FALSE}
options(tinytex.verbose = TRUE)
```

```{r, include = FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
library(sjPlot)
library(ggpubr)
```

```{r, include=FALSE}
# Goodness of Fit
# THESE FUNCTIONS AVAILABLE IN WEEK12_CLASS R FILE
pois_pearson_gof <-
  function(model) {
    return(
      list(
        pval = tibble(
          pred = predict(model, type = "response"),
          y = model$y
        ) %>%
          {sum((.$y - .$pred)^2/.$pred)} %>%
          pchisq(., model$df.residual, lower.tail = F),
        df = model$df.residual
      )
    )
  }

pois_dev_gof <-
  function(model) {
    return(
      list(
        pval = pchisq(model$deviance, model$df.residual, lower.tail=F),
        df = model$df.residual
      )
    )
  }
glm.RR <- function(GLM.RESULT, digits = 2) {
  
  if (GLM.RESULT$family$family == "binomial") {
    LABEL <- "OR"
  } else if (GLM.RESULT$family$family == "poisson") {
    LABEL <- "RR"
  } else {
    LABEL <- "RR"
  }
  
  COEF      <- stats::coef(GLM.RESULT)
  CONFINT   <- stats::confint(GLM.RESULT)
  TABLE     <- cbind(coef=COEF, CONFINT)
  TABLE.EXP <- round(exp(TABLE), digits)
  
  colnames(TABLE.EXP)[1] <- LABEL
  
  TABLE.EXP
}
```

## __Introduction__

Cardiovascular disease (CVD) is a leading cause of death both in California and the United States.The environment problems and source of pollutions have been shown to increase the risk of CVD rate. PM2.5 and other air pollutants such as ozone, nitrogen dioxide, and sulfur dioxide have been considered in the development of health-based standards. Moreover, people living in communities that were identified as "disadvantage" by California Environmental Protection Agency (CalEPA) are more vulnerable to the effects of pollution than others.

The main purpose of this project is to evaluate the association between pollution burden and CVD rate while accounting for community's vulnerability to this association in all California counties.

## __Methods__

### __Data source__ 
The data set was downloaded from (https://data.ca.gov/dataset/calenviroscreen-3-0-results) through API. Once downloaded, the desired information was extracted and formed into a data table. This is a data set including environmental, health, and socioeconomic information of all communities in California State. The key independent variables that were examined in this study were __PM2.5__, __ozone__, __traffic density__, __cardiovascular disease rate__ and __community category__. 
  
### __Examine variables__
1) __PM2.5__ PM2.5 exposure in this data set was an annual mean concentration(ug/m3). After assessing the linearity of PM2.5 and CVD rate, the association was roughly linear. 
2) __Ozone__ Ozone exposure, which is the mean of summer month (May-October) of the daily maximum 8 hours concentration (ppm). The relationship between ozone exposure and CVD rate was linear.
3) __Traffic density__ Traffic exposure(vehicles-km/hr/km) is represented as the number of vehicles (adjusted by road segment lengths in kilometers) per hour per kilometer of roadways.Traffic variable was highly skewed. In order to easily interpret, the traffic density was cut into four levels, and the likelihood test result suggested that keep traffic density variable categorical was better. 
4) __The events of cardio_disease__ Cardiovascular disease rate is a age-adjusted rate of emergency department(ED) visits for AMI per 10,000 residents. The events of cardio-disease were calculated through multiplying the rate of ED visit per 10,000 residents by the population in that community. Two missing data from population was excluded. 
5) __Community__ The communities were divided into two categories: disadvantaged and non-disadvantaged. California Environmental Protection Agency (CalEPA) identify the communities which are more vulnerable to the effects of pollution than others as disadvantaged communities.

### __Statistical analysis__
   An modified Poisson regression model was used to estimate the association of pollutants and cardiovascular disease rate in all communities of California State. Analyses were performed by adjusting for community category confounding factor. Further model fit assessment was tested by Person chi-square test. The overdispersion of the  model was also tested to determine if the variance was larger than what would be expected under a Poisson distribution. Negative binomial regression model was used to address overdispersion issue. P-values< 0.05 were considered statistically significant.
   
```{r,include = FALSE} 
# load the data set by API
library(httr)
go_query <- GET("https://data.ca.gov/api/3/action/datastore_search?resource_id=89b3f4e9-0bf8-4690-8c6f-715a717f3fae&limit=10000")
dat <- content(go_query)
dat <- as.list(dat)
dat_1 <- do.call(rbind,lapply(dat$result$records, rbind))
dat_1[dat_1 == "NULL"]= NA
data <- as.data.frame(dat_1) 
data <- as.data.table(lapply(data, unlist))
data <- as.data.table(lapply(data,function(x) type.convert(as.character(x), as.is = TRUE)))
## create a new data set including key variables
data1 <-
  data %>% 
  select(`Total Population`,ZIP, `California County`,Latitude, Longitude,`SB 535 Disadvantaged Community`,Ozone,`Ozone Pctl`,PM2.5, `PM2.5 Pctl`,Traffic,`Traffic Pctl`, `Cardiovascular Disease`,Asthma,Education, Poverty,Unemployment) %>% 
  as.data.table()
#rename column 
colnames(data1) = c("pop_unit","ZIP","county","latitude","longitude","community_cat", "ozone","ozone_pct","PM2.5","PM2.5_pct","traffic",
                    "traffic_pct","cardio","asthma","education","poverty","unemployment")
# calculate the mean value of key variables
data1[, PM2.5_mean := mean(PM2.5, na.rm = TRUE), by = county]
data1[, ozone_mean := mean(ozone, na.rm = TRUE), by = county]
data1[, traffic_mean := mean(traffic, na.rm = TRUE), by = county]
data1[, cardio_mean := mean(cardio, na.rm = TRUE), by = county]

# create a new variable for pollution burden score.Pollution Burden is calculated as the average of pollution exposures percentile (Ozone_Pct,PM2.5_Pct,Traffic_Pct,tox_Pct)
data1[, pol_score := (ozone_pct+PM2.5_pct+traffic_pct)/3]
data1[, pol_score1 := (pol_score*10)/max(data1$pol_score, na.rm = TRUE)]
data1[, score_mean := mean(pol_score1, na.rm = TRUE), by = county]

# make cardio-disease case numbers and total population
data1 <-
  data1 %>% 
  mutate(community = as.integer(community_cat=="Yes"),
         cardio_count= round((cardio*pop_unit)/10000,digit=0))
# create data2 only has 58 observations
data2<-as.data.table(data1)[, .SD[1], by="county"]
```

<br>

## __Results__

###  __The relationship between pollution particles and CVD rate__
```{r, message=FALSE, include=FALSE}
p1_scatter <- ggplot(data2, aes(x=PM2.5_mean, y=cardio)) + 
  geom_point()+ 
  geom_smooth(method=lm)+
  labs(title = "Pollutant with CVD rate(per 10,000 residents)", x="PM2.5(ug/m3)",y ="Cardiovascular rate")

p2_scatter <- ggplot(data2, aes(x=ozone_mean, y=cardio)) + 
  geom_point()+ 
  geom_smooth(method=lm)+
  labs(x="ozone(ppm)")

p3_scatter<- ggplot(data2, aes(x=score_mean, y=cardio)) + 
  geom_point()+ 
  geom_smooth(method=lm)+
  labs(x="pollution score")
```

```{r, echo=FALSE, message=FALSE}
subplot(p1_scatter,p2_scatter,p3_scatter,nrows=1,shareY=TRUE,titleX = TRUE)
```
__Figure 1__: The correlation between cardiovascular disease (CVD) rate and pollution particle: PM2.5 and ozone, by county.

PM2.5 and ozone, are main particles considered in the study. Pollution score is an averaged pollution burden of environmental effects for each county. It is calculated as the average percentiles,the percentile represents a relative score for the indicator, of the environmental effects indicators. The county with higher score therefore has relatively high pollution burdens. In __Figure 1__, PM2.5 and ozone are positively related with CVD rate. The pollution score plot is also showing a positive correlation with CVD rate.


```{r,map for pollution score and CVD rate, include=FALSE}
 library(urbnmapr)
 library(ggpubr)
#trim white space in county.
county_names <- trimws(data1$county, which = c("both", "left", "right"), whitespace = "[ \t\r\n]")
#append "County" to the trimmed county_names and add it as column "county_name" in data
data1 <-data1 %>% 
  mutate(county_name = paste(county_names, "County", sep = " "))
#construct a new dataset that contains only cardio_mean and county_name
to_map <- data1%>% 
  select(cardio_mean, county_name, pol_score,PM2.5_mean, ozone_mean,traffic_mean)
#select unique county and cadio_mean.
 to_map<-as.data.table(to_map)[, .SD[1], by="county_name"]
#urbanmap data from library(urbnmapr)...
urbnmap_data <- countydata %>% 
  left_join(counties, by = "county_fips") %>% 
  filter(state_name =="California")
#merge to_map with urbanmap_data so that we can draw an urban map of cardio_mean by county.
#name the merged data as new_data
new_data<-merge(x=urbnmap_data, y=to_map, 
      by="county_name",
      all.x = TRUE,
      all.y = FALSE)
#draw polution index map.
g6<- new_data%>% 
  ggplot(mapping = aes(long, lat, group = group, fill=pol_score ))+
  geom_polygon(color = "#ffffff", size = .25)+
  scale_fill_gradientn(colours = c("#CFE8F3", "#A2D4EC", "#73BFE2", "#46ABDB", "#1696D2", "#12719E","#0A4C6A", "#062635"),guide = guide_colorbar(title.position = "top")) +
  theme(legend.title = element_text(),
        legend.key.width = unit(.5, "in")) +
  labs(title=("Pollution score"),
       fill = "Pollution_score")
#draw cardiovascular disease rate map.
g5<- new_data%>% 
  ggplot(mapping = aes(long, lat, group = group, fill=cardio_mean ))+
  geom_polygon(color = "#ffffff", size = .25)+
  scale_fill_gradientn(colours = c("#CFE8F3", "#A2D4EC", "#73BFE2", "#46ABDB", "#1696D2", "#12719E","#0A4C6A", "#062635"),guide = guide_colorbar(title.position = "top")) +
  theme(legend.title = element_text(),
        legend.key.width = unit(.5, "in")) +
  labs(title=("Cardiovascular disease rate"),
       fill = "CVD rate")
       
```

```{r, echo=FALSE, message=FALSE}
map<-ggarrange(g5, g6,
          labels = c("A", "B"),
          ncol = 2, nrow = 2)
 # annotate_figure(map, top = text_grob("Figure2. pollution exposure and CVD rate by county "))  
```

  __Figure 2__: State map depict the pollution score and CVD rate data of each county in California.

Counties with high pollution score is consistent with higher CVD rate. For example, Madera, Fresno, Los Angeles, San Joaquin, San Bernadino and Riverside counties have higher pollution score as well as a high CVD rate( __Figure 2__ ). However, there are some inconsistency. For instance, the northern counties of state shows lower pollution burden but have relatively higher CVD rate. This suggests that other life characteristics and factors may contribute to their high CVD rate.

<br>

### __Pollution burden and community category__
```{r, echo=FALSE, message=FALSE}
library(knitr)
table1 <-
data1 %>% 
  summarise(county,score_mean,community_cat) %>% 
  arrange(desc(score_mean))


# calculate the proportion of disadvantaged communities for each county
table2<-
  data1 %>% 
  group_by(county) %>% 
  summarise(score_mean=mean(score_mean),community_prop = mean(community)) %>% 
  arrange(desc(score_mean))

table3<-head(as.data.table(table2)[,.SD[1],by="county"], 10)
```

```{r,echo=FALSE, message=FALSE}
kable(table3, align = "lcr",
      col.names = c("County","Pollution Score","Disadvantaged Community"))

```
 
```{r,echo=FALSE, message=FALSE}
# bar chart sort by community_prop value
table3$county <- factor(table3$county,levels = unique(table3$county)[order(table3$score_mean,decreasing = TRUE)])

fig<- plot_ly(table3,x = ~county,y = ~community_prop, type='bar',text =~paste(paste(" pollution score: ", score_mean, sep="")))
fig <- fig %>% 
layout(title =("Dis-communities proportion in county"),
                       xaxis =list(title=("County")) ,
                       yaxis = list(title=("Proportion")))

```

```{r, echo=FALSE, warning=FALSE}
fig
```

 __Figure 3__:  __Figure 3__ lists top 10 counties with the higher pollution score. Fresno county has the highest pollution burden and is approximately 60% disadvantaged communities in this county. A high pollution burden is not exactly positive related with high proportion of disadvantaged communities of the county. 

<br>

```{r,include=FALSE, message=FALSE, warning=FALSE}
p<- 
  plot_ly(data1, x = ~as.factor(community_cat), y = ~cardio) %>% 
  add_boxplot(color = ~community_cat) %>%
  layout(title ="CVD rate by community category ",
         legend_title = "dis",
         yaxis = list(title = "cardiovascular rate(per 10,000 residents)"),
         xaxis = list(title = "disadvantaged community")) 
  
```

```{r, echo=FALSE, warning=FALSE}
p
```
__Figure 3__ : CVD rate by community category.  

   Among the top 10 cities with high pollution burden, 90% of them are disadvantaged communities. CalEPA classifies communities as "disadvantaged" due to the environmental conditions and vulnerability of people living in those communities. __Figure3__ and __table1__ both indicate that living in an disadvantaged community increase the risk of CVD rate.

<br>

###  __CVD rate with pollution burden by community category__
```{r,include=FALSE, warning=FALSE}
p <- data1 %>% 
  plot_ly(x = ~score_mean, y = ~cardio_mean,
          type = 'scatter', mode = 'markers', color = ~community_cat,
          size = ~cardio_mean, sizes = c(5, 70), 
          marker = list(sizemode='diameter', opacity=0.5),
          text = ~paste( paste(county, sep=""), paste(" pollution score: ", score_mean, sep="") , paste("cardiovascular rate : ", cardio_mean, sep=""), sep = "<br>")) %>%
  layout(title = "CVD rate vs. pollutant burden by community category",legend_title = "disadvantaged community",
                yaxis = list(title = "Cardiovascular rate(per 10,000 residents)"), 
         xaxis = list(title = "pollution burden"),hovermode = "compare")

```

```{r, echo=FALSE, warning=FALSE}
p
```
__Figure 4__ : Cardiovascular rate vs. pollutant burden by community category.  

__Figure 4__ shows the association between CVD rate and pollution burden by community category. The average pollution burden in disadvantaged  communities is obviously higher than non-disadvantage communities. However, the relationship between CVD rate and pollution burden is not as obvious as in __Figure 1__.
   
<br> 


```{r simple relationship table2, include=FALSE}
# make traffic variable to categorical predictor
data1 <-
  data1 %>%
  mutate(
    traffic.q4 = cut(traffic,
                     breaks = quantile(traffic, probs = 0:4/4,na.rm = T),
                     include.lowest = T),
    traffic.q4_f = as.factor(traffic.q4),
    ozone_ctr= ozone/mean(ozone)
  )
# remove o counts in population
data1 <- data1[!data1$pop_unit==0]
```

### __Traffic density level VS PM2.5 and Ozone__
```{r interaction PM2.5 and traffic,include=FALSE, warning=FALSE}
t3<-
data1 %>% 
  select(traffic.q4,PM2.5_mean,ozone_mean) %>%
  group_by(traffic.q4) %>% 
  summarise(PM2.5 = mean(PM2.5_mean),ozone=mean(ozone_mean))
t3<-t3[-5,]
```

```{r, echo=FALSE}
kable(t3,align = "lcr",
      col.names = c("Traffic density level (vehicles-km/hr/km)","PM2.5 (ug/m3)","Ozone (ppm)"))
```
__Table 3__: Traffic density vs PM2.5 and Ozone

The lowest traffic density level was 22.4-442 (vehicles-km/hr/km).The mean PM2.5 and ozone value associated with the lowest traffic density were 9.98(ug/m3) and 0.049(ppm), respectively. The highest traffic density level was associated the highest PM2.5 value 10.62(ug/m3). Thus, increased PM2.5 was associated with high traffic density level.

<br>

### __Adjusted association between CVD rate and all variables__
```{r final negative model, include=FALSE, warning=FALSE}
#preliminary model:
model1<- glm(cardio_count~PM2.5*traffic.q4+ozone_ctr, family=poisson, data=data1)

#adjust for community :
model_adj<-glm(cardio_count~PM2.5*traffic.q4+ozone_ctr++community, family=poisson, data=data1)
summary(model_adj)

# GOF test : p-value is small, not fit well
pois_pearson_gof(model_adj)
pois_dev_gof(model_adj)

# Checking for Overdispersion
AER::dispersiontest(model_adj)

# negative binomial model
model_neg<- MASS::glm.nb(cardio_count~PM2.5*traffic.q4+ozone_ctr+community, link=log, data=data1)

pois_dev_gof(model_neg)

# table for negative model

q2<-tab_model(model1,model_neg,
              dv.labels = c("Unadjusted Model", "Negative binomial  Model"),
             string.ci = "Conf. Int (95%)",
             string.p = "P-Value",show.reflvl = TRUE,
             prefix.labels = "varname",file = 'fgf.doc'
  )

```

```{r, echo=FALSE}
q2
```
__Table 5__: Adjusted association between the pollution and CVD rate.
   
After checking overdispersion, there was an 1.14 (P<0.001) overdispersion term in the Poisson model. A negative binomial regression model was performed to address overdispersion. In the final negative binomial model, PM2.5 was not statistically associated with CVD rate(p=0.463) compared to unadjusted model(p<0.001), and the exposure had the similar CVD rate as non-exposure(95% CI 0.99,1.00;p=0.463). Every one-unit ozone exposure is associated with 2.21 times the CVD rate(95% CI 2.08,2.35;p<0.001), while 2.27 times before adjusting for community. The highest traffic density exposure was associated with 1.84 times CVD rate (95% CI 1.53,2.21;p<0.001) which increased 25% compared to unadjustment model (95% CI 1.24,1.75;p<0.001). Moreover, there is a significant interaction between PM2.5 and traffic density,and increased PM2.5 was associated with high traffic density level( __table3__ ).
   
<br>

## __Conclusion__

Pollutants such PM2.5, Ozone and traffic were statistically significantly associated with CVD rate and living in an disadvantaged community increases risk of CVD rate. The baseline of CVD rate after adjustment of community category increased 13%. The CVD rate in highest traffic density level increased 25% compared to unadjusted model. Although the CVD rate associated with PM2.5, and ozone exposure did not change much after adjustment, the people with ozone exposure still had 2.21 times the rate of CVD than the baseline. PM2.5 had a significant interaction with traffic density level. Thus, improvements in air quality would be helpful to reduce overall CVD prevalence across California. In addition, efforts to identify the pollution source that accounting for a community's vulnerability would beneficial to those disadvantaged communities which suffered most from CVD occurrence. 

<br>
<br>


