---
title: "Association of air pollution and cardiovascular disease rate in california counties"
author: "Luqing Ren"
always_allow_html: TRUE
output:
  html_document:
      toc: TRUE
      toc_float: TRUE
---


```{r, include = FALSE}
options(tinytex.verbose = TRUE)
```

```{r, include = FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
library(sjPlot)
```

```{r, include=FALSE}
# Goodness of Fit
# THESE FUNCTIONS AVAILABLE IN WEEK12_CLASS R FILE
pois_pearson_gof <-
  function(model) {
    return(
      list(
        pval = tibble(
          pred = predict(model, type = "response"),
          y = model$y
        ) %>%
          {sum((.$y - .$pred)^2/.$pred)} %>%
          pchisq(., model$df.residual, lower.tail = F),
        df = model$df.residual
      )
    )
  }

pois_dev_gof <-
  function(model) {
    return(
      list(
        pval = pchisq(model$deviance, model$df.residual, lower.tail=F),
        df = model$df.residual
      )
    )
  }
glm.RR <- function(GLM.RESULT, digits = 2) {
  
  if (GLM.RESULT$family$family == "binomial") {
    LABEL <- "OR"
  } else if (GLM.RESULT$family$family == "poisson") {
    LABEL <- "RR"
  } else {
    LABEL <- "RR"
  }
  
  COEF      <- stats::coef(GLM.RESULT)
  CONFINT   <- stats::confint(GLM.RESULT)
  TABLE     <- cbind(coef=COEF, CONFINT)
  TABLE.EXP <- round(exp(TABLE), digits)
  
  colnames(TABLE.EXP)[1] <- LABEL
  
  TABLE.EXP
}
```

## __Introduction__

Cardiovascular disease (CVD) is a leading cause of death both in California and the United States.The environment problems and source of pollutions have been shown to increase the risk of CVD rate. PM2.5 and other air pollutants such as ozone, nitrogen dioxide, and sulfur dioxide have been considered in the development of health-based standards. Moreover, people living in communities that were identified as "disadvantage" by California Environmental Protection Agency (CalEPA) are more vulnerable to the effects of pollution than others.

The main purpose of this project is to evaluate the association between pollution burden and CVD rate while accounting for community's vulnerability to this association in all California counties.

## __Methods__

### __Data source__ 
The data set was downloaded from (https://data.ca.gov/dataset/calenviroscreen-3-0-results) through an API. Once downloaded, the desired information was extracted and formed into a data table. This is a data set including environmental, health, and socioeconomic information of all communities in California State. The key independent variables that were examined in this study were __PM2.5__, __ozone__, __traffic density__, __cardiovascular disease rate__ and __community category__. 
  

### __Statistical analysis__
   An modified Poisson regression model was used to estimate the association of pollutants and cardiovascular disease rate in all communities of California State. Analyses were performed by adjusting for community category confounder. Further model fit assessment was tested by Person chi-square test. The overdispersion of the  model was also tested to determine if the variance was larger than what would be expected under a Poisson distribution.Negative binomial regression model was used to address overdispersion. P-values< 0.05 were considered statistically significant.
   
```{r,include = FALSE} 
# load the data set by API
library(httr)
go_query <- GET("https://data.ca.gov/api/3/action/datastore_search?resource_id=89b3f4e9-0bf8-4690-8c6f-715a717f3fae&limit=10000")
dat <- content(go_query)
dat <- as.list(dat)
dat_1 <- do.call(rbind,lapply(dat$result$records, rbind))
dat_1[dat_1 == "NULL"]= NA
data <- as.data.frame(dat_1) 
data <- as.data.table(lapply(data, unlist))
data <- as.data.table(lapply(data,function(x) type.convert(as.character(x), as.is = TRUE)))
## create a new data set including key variables
data1 <-
  data %>% 
  select(`Total Population`,ZIP, `California County`,Latitude, Longitude,`SB 535 Disadvantaged Community`,Ozone,`Ozone Pctl`,PM2.5, `PM2.5 Pctl`,Traffic,`Traffic Pctl`, `Cardiovascular Disease`,Asthma,Education, Poverty,Unemployment) %>% 
  as.data.table()
#rename column 
colnames(data1) = c("pop_unit","ZIP","county","latitude","longitude","community_cat", "ozone","ozone_pct","PM2.5","PM2.5_pct","traffic",
                    "traffic_pct","cardio","asthma","education","poverty","unemployment")
# calculate the mean value of key variables
data1[, PM2.5_mean := mean(PM2.5, na.rm = TRUE), by = county]
data1[, ozone_mean := mean(ozone, na.rm = TRUE), by = county]
data1[, traffic_mean := mean(traffic, na.rm = TRUE), by = county]
data1[, cardio_mean := mean(cardio, na.rm = TRUE), by = county]

# create a new variable for pollution burden score.Pollution Burden is calculated as the average of pollution exposures percentile (Ozone_Pct,PM2.5_Pct,Traffic_Pct,tox_Pct)
data1[, pol_score := (ozone_pct+PM2.5_pct+traffic_pct)/3]
data1[, pol_score1 := (pol_score*10)/max(data1$pol_score, na.rm = TRUE)]
data1[, score_mean := mean(pol_score1, na.rm = TRUE), by = county]

# make cardio-disease case numbers and total population
data1 <-
  data1 %>% 
  mutate(community = as.integer(community_cat=="Yes"),
         cardio_count= round((cardio*pop_unit)/10000,digit=0))
# creat data2 only has 58 observations
data2<-as.data.table(data1)[, .SD[1], by="county"]
```

<br>

## __Results__

###  __The relationship between pollution particles and CVD rate__
```{r, message=FALSE, include=FALSE}
p1_scatter <- ggplot(data2, aes(x=PM2.5_mean, y=cardio)) + 
  geom_point()+ 
  geom_smooth(method = "lm")+
  labs(title = "Pollutant with CVD rate", x="PM2.5",y ="Cardiovascular rate")

p2_scatter <- ggplot(data2, aes(x=ozone_mean, y=cardio)) + 
  geom_point()+ 
  geom_smooth(method = "lm")+
  labs(x="ozone")

p3_scatter<- ggplot(data2, aes(x=score_mean, y=cardio)) + 
  geom_point()+ 
  geom_smooth(method = "lm")+
  labs(x="pollution score")
```

```{r, echo=FALSE, message=FALSE}
subplot(p1_scatter,p2_scatter,p3_scatter,nrows=1,shareY=TRUE,titleX = TRUE)
```
__Figure 1__: The correlation between cardiovascular disease (CVD) rate and pollution particle: PM2.5 and ozone, by county.

PM2.5 and ozone, are main particles considered in the study. Pollution score is an averaged pollution burden of environmental effects for each county. It is calculated as the average percentiles,the percentile represents a relative score for the indicator, of the environmental effects indicators. The county with higher score therefore has relatively high pollution burdens. In __Figure 1__, PM2.5 and ozone are positively related with CVD rate. The pollution score plot is also showing a positive correlation with CVD rate.


```{r,map for pollution score and CVD rate, include=FALSE}
 library(urbnmapr)
 library(ggpubr)
#trim white space in county.
county_names <- trimws(data1$county, which = c("both", "left", "right"), whitespace = "[ \t\r\n]")
#append "County" to the trimmed county_names and add it as column "county_name" in data
data1 <-data1 %>% 
  mutate(county_name = paste(county_names, "County", sep = " "))
#construct a new dataset that contains only cardio_mean and county_name
to_map <- data1%>% 
  select(cardio_mean, county_name, pol_score,PM2.5_mean, ozone_mean,traffic_mean)
#select unique county and cadio_mean.
 to_map<-as.data.table(to_map)[, .SD[1], by="county_name"]
#urbanmap data from library(urbnmapr)...
urbnmap_data <- countydata %>% 
  left_join(counties, by = "county_fips") %>% 
  filter(state_name =="California")
#merge to_map with urbanmap_data so that we can draw an urban map of cardio_mean by county.
#name the merged data as new_data
new_data<-merge(x=urbnmap_data, y=to_map, 
      by="county_name",
      all.x = TRUE,
      all.y = FALSE)
#draw polution index map.
g6<- new_data%>% 
  ggplot(mapping = aes(long, lat, group = group, fill=pol_score ))+
  geom_polygon(color = "#ffffff", size = .25)+
  scale_fill_gradientn(colours = c("#CFE8F3", "#A2D4EC", "#73BFE2", "#46ABDB", "#1696D2", "#12719E","#0A4C6A", "#062635"),guide = guide_colorbar(title.position = "top")) +
  theme(legend.title = element_text(),
        legend.key.width = unit(.5, "in")) +
  labs(title=("Pollution score"),
       fill = "Pollution_score")
#draw cardiovascular disease rate map.
g5<- new_data%>% 
  ggplot(mapping = aes(long, lat, group = group, fill=cardio_mean ))+
  geom_polygon(color = "#ffffff", size = .25)+
  scale_fill_gradientn(colours = c("#CFE8F3", "#A2D4EC", "#73BFE2", "#46ABDB", "#1696D2", "#12719E","#0A4C6A", "#062635"),guide = guide_colorbar(title.position = "top")) +
  theme(legend.title = element_text(),
        legend.key.width = unit(.5, "in")) +
  labs(title=("Cardiovascular disease rate"),
       fill = "CVD rate")
       
```

```{r, echo=FALSE, message=FALSE}
map<-ggarrange(g5, g6,
          labels = c("A", "B"),
          ncol = 2, nrow = 2)
annotate_figure(map, top = text_grob("Figure2. pollution exposure and CVD rate by county "))  
```

  __Figure 2__: State map depict the pollution score and CVD rate data of each county in California.

Counties with high pollution score is consistent with higher CVD rate. For example, Madera, Fresno, Los Angeles, San Joaquin, San Bernadino and Riverside counties have higher pollution score as well as a high CVD rate( __Figure 2__ ). However, there are some inconsistency. For instance, the northern counties of state shows lower pollution burden but have relatively higher CVD rate. This suggests that other life characteristics and factors may contribute to their high CVD rate.

<br>

### __Pollution burden and community category__
```{r, include=FALSE}
library(knitr)
data1<-
data1 %>% 
mutate(score_mean=as.numeric(format(round(data1$score_mean, 3), nsmall = 3)),
       cardio_mean=as.numeric(format(round(data1$cardio_mean, 3), nsmall = 3)))
table1 <-
data1 %>% 
  summarise(county,score_mean,community_cat) %>% 
  arrange(desc(score_mean))
table1<-head(as.data.table(table1)[,.SD[1],by="county"], 10)

```

```{r, echo=FALSE, message=FALSE}
kable(table1, align = "lcr",
      col.names = c("County","Score","Community category"))
```
__Table 1__: Top 10 counties with high pollution burden and their community category. 

<br>

```{r,include=FALSE, message=FALSE, warning=FALSE}
p<- 
  plot_ly(data1, x = ~as.factor(community_cat), y = ~cardio,alpha = 0.1) %>% 
  add_boxplot(color = ~community_cat) %>%
  layout(title ="cardiovascular rate by community category ",
         legend_title = "dis",
         yaxis = list(title = "cardiovascular rate"),
         xaxis = list(title = "disadvantage community"))

```

```{r, echo=FALSE, warning=FALSE}
p
```
__Figure 3__ : CVD rate by community category.  

   Among the top 10 cities with high pollution burden, 90% of them are indisadvantage communities. CalEPA classifies communities as "disadvantage" due to the environmental conditions and vulnerability of people living in those communities. __Figure3__ and __table1__ both indicate that living in an disadvantageous community may increase the risk of CVD.

<br>

###  __CVD rate with pollution burden by community category__
```{r,include=FALSE, warning=FALSE}

p <- data1 %>% 
  plot_ly(x = ~score_mean, y = ~cardio_mean,
          type = 'scatter', mode = 'markers', color = ~community_cat,
          size = ~cardio_mean, sizes = c(5, 70), 
          marker = list(sizemode='diameter', opacity=0.5),
          text = ~paste( paste(county, sep=""), paste(" pollution score: ", score_mean, sep="") , paste("cardiovascular rate : ", cardio_mean, sep=""), sep = "<br>")) %>%
  layout(title = "Cardiovascular rate vs. pollutant burden by community category",legend_title = "disadvantage community",
                yaxis = list(title = "Cardiovascular rate"), 
         xaxis = list(title = "pollution burden"),hovermode = "compare")

```

```{r, echo=FALSE, warning=FALSE}
ggplotly(p)
```
__Figure 4__ : Cardiovascular rate vs. pollutant burden by community category.  

__Figure 4__ shows the association between CVD rate and pollution burden by community category. The average pollution burden in disadvantage communities is obviously higher than non-disadvantage communities. However, the relationship between CVD rate and pollution burden is not as obvious as in __Figure 1__.
   
<br> 


## __Conclusion__

Pollutants such PM2.5, Ozone and traffic were statistically significantly assocaited with CVD rate and living in an disadvantageous community may increases risk of CVD rate. The baseline of CVD rate after adjustment of community category increased 13%. The CVD rate in highest traffic density level increased 25% compared to unadjustment. Although the CVD rate associated with PM2.5, and ozone exposure did not change much after adjustment, the people with ozone exposure still had 2.21 times the rate of CVD than the baseline. PM2.5 had a significant interaction with traffic density level. Thus, improvements in air quality would be helpfull to reduce over all CVD prevalance acrross california. In additon, efforts to identify the pollution source that accounting for a community's vulnerability would benefitial to those disadvantaged communities which suffered most from CVD occurence. 
  
<br>
<br>


